---
title: "Variable selection procedure"
author: "Stephanie Herman"
date: "`r format(Sys.time())`"
output: 
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: spacelab
    highlight: textmate
    df_print: paged
    code_folding: hide
    self_contained: false
    keep_md: false
    encoding: "UTF-8"
    css: ["assets/lab.css"]
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE)
library(here)
library(glmnet)
library(ggplot2)
library(tidyr)
library(pROC)
library(reshape2)
library(gridExtra)
library(knitr)
library(gtools)
library(caret)
library(dplyr)
library(ggpubr)
library(viridis)
library(emmeans)
library(RColorBrewer)
```


```{r, include=FALSE}
# load functions
source(file.path(here(), 'data_processing/Functions.R'))
set.seed(435)

# load matching
load(file.path(here(), 'data/MS3data.RData'))
load(file.path(here(), 'data/MS1data.RData'))

# load matched features
load('featureMatch.RData')
```


```{r}
# save untouched data for later
MS3data$untouched <- MS3data$data
MS1data$untouched <- MS1data$data

# format data for analysis
MS3data$data <- data.frame(t(MS3data$data[, -c(1,2)]), stringsAsFactors = F)
MS1data$data <- data.frame(t(MS1data$data[, -c(1,2)]), stringsAsFactors = F)
```

```{r}
cutoff <- .9

# coverage cutoff MS3
ind <- which(apply(MS3data$data, 2, function(x) sum(!is.na(x))/length(x))>=cutoff)
MS3data$data <- MS3data$data[, ind]

# coverage cutoff MS1
ind <- which(apply(MS1data$data, 2, function(x) sum(!is.na(x))/length(x))>=cutoff)
MS1data$data <- MS1data$data[, ind]
```

```{r, include = T}
# extract age dependent metabos
ind <- which(MS3data$metadata$Condition == 'ctrl')

pVal <- c()
for (i in 1:ncol(MS3data$data)) {
  c <- cor.test(as.numeric(MS3data$data[ind,i]), MS3data$metadata$Age[ind], na.action = "na.omit")
  pVal <- c(pVal, c$p.value)
}
n <- which(pVal<0.05)
age.rm <- names(MS3data$data)[n]
MS3data$data <- MS3data$data[, -n]

#cat('Number of removed age dependent metabolites: ', length(age.rm))
```

# Linking the cohorts

The features were matched between cohorts using a **±2.5 ppm** and **±5 sec**. Additionally, an average intensity deviation less than **±5 standard deviation** was allowed and only metabolic features present in 90% of the samples in both cohorts were kept. Finally, metabolic features that show a statistically significant (p-value<0.05) age dependence in the healthy control subjects in cohort 1 were excluded. 

Remaining missing values were replaced by the average feature value and features were centered separately within both cohorts.

```{r, include = T, echo = T}
# rename MS1 metabolites
row.names(featureMatch) <- featureMatch$MS1
names(MS1data$data) <- paste0('X', featureMatch[gsub('X', '', names(MS1data$data)), 'MS3'])

common <- intersect(names(MS1data$data), names(MS3data$data))
MS1data$data <- MS1data$data[, common]
MS3data$data <- MS3data$data[, common]

# intensity diff cutoff
int.MS3 <- apply(MS3data$data[which(MS3data$metadata$Collection == 1), ], 2, function(x) mean(x, na.rm=T))
int.MS1 <- apply(MS1data$data, 2, function(x) mean(x, na.rm=T))

intDiff <- apply(cbind(int.MS3, int.MS1), 1, function(x) x[1]-x[2])
ind <- which(abs(intDiff)<sd(intDiff))

MS1data$data <- MS1data$data[, ind]
MS3data$data <- MS3data$data[, ind]

cat('Number of final metabolic features: ', length(ind))

# impute by average
colmissing <- which(apply(is.na(MS3data$data), 2, any))
for (i in 1:length(colmissing)) {
  temp <- MS3data$data[,colmissing[i]]
  temp[is.na(temp)] <- mean(temp, na.rm=T)
  MS3data$data[,colmissing[i]] <- temp
}

colmissing <- which(apply(is.na(MS1data$data), 2, any))
for (i in 1:length(colmissing)) {
  temp <- MS1data$data[,colmissing[i]]
  temp[is.na(temp)] <- mean(temp, na.rm=T)
  MS1data$data[,colmissing[i]] <- temp
}

# center features within cohort
MS3data$data <- scale(MS3data$data, center = T, scale = F)
MS1data$data <- scale(MS1data$data, center = T, scale = F)

# extract RRMS and PMS patients, first samplings, MS3 cohort
MS3meta <- MS3data$metadata[MS3data$metadata$Collection==1,]
MS3meta <- MS3meta[which(MS3meta$Condition=='PMS' | MS3meta$Condition=='RRMS'),]
MS3train <- data.frame(MS3data$data[MS3meta$Code,])
MS3train <- cbind(MS3meta$Condition, MS3train)
names(MS3train)[1] <- 'Condition'

# extract RRMS and PMS patients, MS1 cohort
MS1meta <- MS1data$metadata[which(MS1data$metadata$Condition=='PMS' | MS1data$metadata$Condition=='RRMS'), ]
MS1train <- data.frame(MS1data$data[row.names(MS1meta),])
MS1train <- cbind(MS1meta$Condition, MS1train)
names(MS1train)[1] <- 'Condition'
```

# Train GLMNET model

Variable selection was done solely on the training (MS3) cohort. The training cohort consisted of **35 RRMS** and **35 PMS** patients. A regularized regression with an alpha = 0.5 (elastic-net regularization) was trained with a lambda of 1 standard error from the lambda that minimized the cross validated deviance in a balanced 7-fold cross validation.

```{r, include = T, echo = T}
# scale training data
X <- scale(MS3train[, -1], center = T, scale = T)
train_scaling <- list()
train_scaling$center <- attr(X, "scaled:center")
train_scaling$scale <- attr(X, "scaled:scale")
Y <- as.character(MS3train$Condition)
X <- data.frame(X)
train <- cbind(Y, X)
train <- train[sort(as.character(train$Y), decreasing = T, index.return = T)$ix, ]

alpha.fit <- .5
foldid <- c(rep(sample.int(7),5), rep(sample.int(7),5))

# optimize lambda
elm_fit <- cv.glmnet(as.matrix(train[, -1]), 
                     factor(train$Y, levels = c('RRMS', 'PMS')), 
                     family="binomial", 
                     alpha=alpha.fit, 
                     standardize = F, 
                     foldid = foldid)

# train model with optimized lambda
model <- glmnet(as.matrix(train[, -1]), 
                factor(train$Y, levels = c('RRMS', 'PMS')), 
                family="binomial", 
                lambda = elm_fit$lambda.1se, 
                alpha = alpha.fit, 
                standardize = F)
```

```{r, include = T, fig.height = 6, fig.width = 5, fig.align = 'center'}
# extract model coefficients
coefs <- model$beta[, 1]
coefs <- data.frame(Coefficient = coefs[-which(coefs == 0)])
coefs$Metabo <- row.names(coefs)
cat('Number of selected metabolic features: ', nrow(coefs))

coefs <- coefs[sort(abs(coefs$Coefficient), decreasing = T, index.return = T)$ix, ]
coefs$Metabo <- factor(coefs$Metabo, levels = coefs$Metabo)

cplot <- ggplot(coefs, aes(x = Metabo, y = Coefficient)) + 
  geom_bar(stat = 'identity', width = .6, col = 'white', fill = 'darkred') + coord_flip() + 
  theme(legend.position = 'none') + ylab('') + xlab('') + theme_pubr() + geom_hline(yintercept = 0)
cplot
```
**Figure 1.** The selected metabolic features ranked according to their coefficients, $\beta$, from the regularized regression.

# Predict the test cohort

To evaluate the performance of the regularized model, cohort 2 (MS1) was used as an independent test set.

```{r, include = T, echo = T, fig.height = 5, fig.width = 5, fig.align = 'center'}
# scale MS1 cohort
MS1_X <- data.frame(scale(MS1train[, -1], center = train_scaling$center, scale = train_scaling$scale))
MS1_Y <- factor(MS1train$Condition, levels = c('RRMS', 'PMS'))

# predict response for test cohort
prediction <- predict(model, as.matrix(MS1_X), type = "response")
myroc <- roc(MS1_Y, prediction[, 1],
             ci=TRUE, stratified=FALSE,
             plot=FALSE, 
             print.auc=TRUE, show.thres=TRUE)
```

```{r, echo = F, warning = F}
# compute and compare AUC for model vs the selected features in isolation
theroc  <- setNames(data.frame(cbind(1 - rev(myroc$specificities),
                                     rev(myroc$sensitivities))), c('x', 'y'))

metauc <- c()
plotx <- list()
ploty <- list()
n <- 1
for (i in 1:ncol(MS1_X[, as.character(coefs$Metabo)])) {
  metroc <- roc(MS1_Y, MS1_X[, as.character(coefs$Metabo)[i]],
                # arguments for ci
                ci=TRUE, stratified=FALSE,
                # arguments for plot
                print.auc=TRUE, show.thres=TRUE)
  
  metauc <- c(metauc, metroc$auc)
  
  plotx[[n]] <- 1 - rev(metroc$specificities)
  ploty[[n]] <- rev(metroc$sensitivities)
  n <- n + 1
}

x <- melt(plotx)
y <- melt(ploty)
names(x) <- c("x", "remove")
names(y) <- c("y", "Metabolite")
plotroc <- cbind(x,y)
plotroc <- plotroc[,-2]

interval = 0.2
breaks = seq(0, 1, interval)
```

```{r, include = T, fig.height = 5, fig.width = 5, fig.align = 'center'}
plotroc$Metabolite <- factor(plotroc$Metabolite)
mypal <- plasma(length(coefs$Metabo)+1)
roc_plot <- ggplot(plotroc, aes(x = x, y = y, col = Metabolite)) + geom_line(alpha = .4) +
  scale_color_manual(values = mypal) +
  geom_segment(aes(x = 0, y = 0, xend = 1,yend = 1), alpha = 0.9, linetype=2, col = 'gray30') +
  scale_x_continuous(name = "1-Specificity",limits = c(0,1), breaks = breaks, expand = c(0.001,0.001)) +
  scale_y_continuous(name = "Sensitivity", limits = c(0,1), breaks = breaks, expand = c(0.001, 0.001)) +
  theme_bw() +
  theme(axis.ticks = element_line(color = "grey80"), legend.position = "none", text = element_text(size=12)) +
  coord_equal() +
  geom_line(data = theroc, aes(x = x, y = y), col = 'black', size = 1) +
  annotate("text", x = 0.65, y = 0.10, vjust = 0, 
           label = paste0("AUC = ", round(myroc$auc, digits = 2),
                          ' (', round(myroc$ci[1], digits = 2), ' - ', 
                          round(myroc$ci[3], digits = 2), ')'), size = 5) +
  annotate("text", x = 0.65, y = 0.05, vjust = 0, 
           label = paste0("AUC = (", round(range(metauc)[1], digits = 2), ' - ',
                          round(range(metauc)[2], digits = 2), ')'), size = 5) 
roc_plot
```

```{r, include = T, fig.height = 5, fig.width = 5, fig.align = 'center'}
# predict class labels for test cohort
prediction <- predict(model, as.matrix(MS1_X), type = "class")
confusionMatrix(factor(prediction, levels = c('RRMS', 'PMS')), MS1_Y, positive = 'PMS')
```

```{r, include = T, warning = F, fig.align = 'center'}
# plot confusion matrix
cm <- confusionMatrix(factor(prediction, levels = c('RRMS', 'PMS')), MS1_Y, positive = 'PMS')
fourfoldplot(cm$table)
```

# Predict transitioners

```{r, include = T}
# extract and scale transitioners
trans <- rbind(MS3data$data[which(MS3data$metadata$Condition == 'Transition'), ],
               MS1data$data[which(MS1data$metadata$Condition == 'Transition'), ])
trans <- data.frame(scale(trans, center = train_scaling$center, scale = train_scaling$scale))

# predict phenotypes of transitioners
prob <- predict(model, as.matrix(trans), type = "response")
class <- predict(model, as.matrix(trans), type = "class")

print(cbind(prob, class[, 1]))
```

# PCA on both cohorts

To visualize the selected features in the two cohorts, a principal component analysis (PCA) was done on each cohort and the selected features.

```{r}
# PCA on MS3
pca <- prcomp(train[, as.character(coefs$Metabo)], center = F, scale. = F)

vars<- pca$sdev^2
vars<- vars/sum(vars)

plotdata <- data.frame(cbind(pca$x[,1:2], 
                             row.names(train),
                             as.character(train$Y)), 
                       stringsAsFactors = F)
names(plotdata)[3] <- 'Sample.ID'
names(plotdata)[4] <- "Condition"

plotdata$PC1 <- as.numeric(as.character(plotdata$PC1))
plotdata$PC2 <- as.numeric(as.character(plotdata$PC2))

plotdata$Condition <- factor(plotdata$Condition, levels = c('RRMS', 'PMS'))
pca.MS3 <- ggplot(plotdata, aes(x=PC1, y=PC2, col = Condition, shape = Condition, label = Sample.ID)) + 
  geom_point(size = 2) + 
  xlab(paste0('PC1: explaining ',round(vars[1], digits = 3)*100, '% of the variance')) + 
  ylab(paste0('PC2: explaining ',round(vars[2], digits = 3)*100, '% of the variance')) + theme_bw()  +
  scale_shape_manual(values = c(19, 17)) +
  scale_color_manual(values = c('#3C568E', '#CC3333', '#CC3333')) +
  ggtitle('Training (MS3) cohort')

# coded according to age and sex
plotdata.meta <- cbind(plotdata, MS3data$metadata[plotdata$Sample.ID, c('Gender', 'Age')])

pca.MS3.sex <- ggplot(plotdata.meta, aes(x=PC1, y=PC2, col = Gender, shape = Condition, label = Sample.ID)) + 
  geom_point(size = 2) + 
  xlab(paste0('PC1 (',round(vars[1], digits = 3)*100, '%)')) + 
  ylab(paste0('PC2 (',round(vars[2], digits = 3)*100, '%)')) + theme_bw()  +
  scale_shape_manual(values = c(19, 17)) +
  ggtitle('MS3 sexes')

pca.MS3.age <- ggplot(plotdata.meta, aes(x=PC1, y=PC2, col = Age, shape = Condition, label = Sample.ID)) + 
  geom_point(size = 2) + 
  xlab(paste0('PC1 (',round(vars[1], digits = 3)*100, '%)')) + 
  ylab(paste0('PC2 (',round(vars[2], digits = 3)*100, '%)')) + theme_bw()  +
  scale_shape_manual(values = c(19, 17)) +
  ggtitle('MS3 ages') +
  scale_color_viridis(option = 'D', direction = -1) 

# scale all MS3 data
MS3subset <- data.frame(cbind(MS3data$metadata$Condition), 
                        MS3data$data,
                        stringsAsFactors = F)
names(MS3subset)[1] <- 'Condition'
MS3subset[, -1] <- data.frame(scale(MS3subset[, -1], center = train_scaling$center, scale = train_scaling$scale))

# project the controls
MS3ctrl <- MS3subset[MS3subset$Condition == 'ctrl', c('Condition', as.character(coefs$Metabo))]
MS3ctrl <- data.frame(as.matrix(MS3ctrl[, -1]) %*% pca$rotation )
MS3ctrl <- cbind(MS3ctrl[,1:2], row.names(MS3ctrl))
names(MS3ctrl)[3] <- 'Sample.ID'
MS3ctrl$Condition <- 'Control'

plotdata.ctrl <- rbind(plotdata, MS3ctrl)
pca.MS3.ctrl <- ggplot(plotdata.ctrl, aes(x=PC1, y=PC2, col = Condition, shape = Condition, label = Sample.ID, alpha = Condition)) + 
  geom_point(size = 2) + 
  xlab(paste0('PC1: explaining ',round(vars[1], digits = 3)*100, '% of the variance')) + 
  ylab(paste0('PC2: explaining ',round(vars[2], digits = 3)*100, '% of the variance')) + theme_bw()  +
  scale_shape_manual(values = c(15, 19, 17)) +
  scale_color_manual(values = c('black', '#3C568E', '#CC3333')) +
  scale_alpha_manual(values = c(1, .5, .5)) +
  ggtitle('Projected controls')

# project transitioners
MS3trans <- MS3subset[MS3subset$Condition == 'Transition', c('Condition', as.character(coefs$Metabo))]
MS3trans <- data.frame(as.matrix(MS3trans[, -1]) %*% pca$rotation )
MS3trans <- cbind(MS3trans[,1:2], row.names(MS3trans))
names(MS3trans)[3] <- 'Sample.ID'
MS3trans$Condition <- 'Transition'

plotdata.trans <- rbind(plotdata, MS3trans)
pca.MS3.trans <- ggplot(plotdata.trans, aes(x=PC1, y=PC2, col = Condition, shape = Condition, label = Sample.ID, alpha = Condition)) + 
  geom_point(size = 2) + 
  xlab(paste0('PC1: explaining ',round(vars[1], digits = 3)*100, '% of the variance')) + 
  ylab(paste0('PC2: explaining ',round(vars[2], digits = 3)*100, '% of the variance')) + theme_bw()  +
  scale_shape_manual(values = c(19, 17, 8)) +
  scale_color_manual(values = c('#3C568E', '#CC3333', 'black')) +
  scale_alpha_manual(values = c(.5, .5, 1)) +
  ggtitle('Projected transitioners')

# project the follow-up samples from the clinical trial
MS3clin <- MS3subset[grep('T', row.names(MS3subset)), c('Condition', as.character(coefs$Metabo))]
MS3clin <- data.frame(as.matrix(MS3clin[-grep('T1', row.names(MS3clin)), -1]) %*% pca$rotation )
MS3clin <- cbind(MS3clin[,1:2], row.names(MS3clin))
names(MS3clin)[3] <- 'Sample.ID'
MS3clin$Condition <- substrRight(as.character(MS3clin$Sample.ID), 2)

plotdata.time <- rbind(plotdata.ctrl, MS3clin)
plotdata.time$Condition <- factor(plotdata.time$Condition, levels = rev(c('PMS', paste0('T', 2:4), 'RRMS', 'Control')))
bplot <- ggplot(plotdata.time, aes(x=Condition, y=PC1, fill = Condition, label = Sample.ID)) + 
  geom_boxplot(width=0.5, outlier.color = 'white', outlier.size = 0) +
  geom_jitter(width = 0.1, alpha=I(0.5)) + theme_pubr() +
  scale_fill_manual(values = rev(c('#CC3333', '#F36F56', '#F8A88D', '#FDDFD1', '#3C568E', 'white'))) + 
  theme(legend.position = 'none', text = element_text(size=12),
        axis.text = element_text(size = 9.5)) + xlab('') + 
  ylab(paste0('PC1: explaining ',round(vars[1], digits = 3)*100, '% of the variance')) + coord_flip()

# PCA on MS1
pca <- prcomp(MS1_X[, as.character(coefs$Metabo)], center = F, scale. = F)

vars<- pca$sdev^2
vars<- vars/sum(vars)

plotdata <- data.frame(cbind(pca$x[,1:2], 
                             row.names(MS1_X),
                             as.character(MS1_Y)), 
                       stringsAsFactors = F)
names(plotdata)[3] <- 'Sample.ID'
names(plotdata)[4] <- "Condition"

plotdata$PC1 <- as.numeric(as.character(plotdata$PC1))
plotdata$PC2 <- as.numeric(as.character(plotdata$PC2))

plotdata$Condition <- factor(plotdata$Condition, levels = c('RRMS', 'PMS'))
pca.MS1 <- ggplot(plotdata, aes(x=PC1, y=PC2, col = Condition, shape = Condition, label = Sample.ID)) + 
  geom_point(size = 2) + 
  xlab(paste0('PC1: explaining ',round(vars[1], digits = 3)*100, '% of the variance')) + 
  ylab(paste0('PC2: explaining ',round(vars[2], digits = 3)*100, '% of the variance')) + theme_bw()  +
  scale_shape_manual(values = c(19, 17)) +
  scale_color_manual(values = c('#3C568E', '#CC3333', '#CC3333')) +
  ggtitle('Test (MS1) cohort')

# coded according to age and sex
plotdata.meta <- cbind(plotdata, MS1data$metadata[plotdata$Sample.ID, c('Gender', 'Age')])

pca.MS1.sex <- ggplot(plotdata.meta, aes(x=PC1, y=PC2, col = Gender, shape = Condition, label = Sample.ID)) + 
  geom_point(size = 2) + 
  xlab(paste0('PC1 (',round(vars[1], digits = 3)*100, '%)')) + 
  ylab(paste0('PC2 (',round(vars[2], digits = 3)*100, '%)')) + theme_bw()  +
  scale_shape_manual(values = c(19, 17)) +
  ggtitle('MS1 sexes')

pca.MS1.age <- ggplot(plotdata.meta, aes(x=PC1, y=PC2, col = Age, shape = Condition, label = Sample.ID)) + 
  geom_point(size = 2) + 
  xlab(paste0('PC1 (',round(vars[1], digits = 3)*100, '%)')) + 
  ylab(paste0('PC2 (',round(vars[2], digits = 3)*100, '%)')) + theme_bw()  +
  scale_shape_manual(values = c(19, 17)) +
  ggtitle('MS1 ages') +
  scale_color_viridis(option = 'D', direction = -1) 

# scale all MS1 data
MS1subset <- data.frame(cbind(MS1data$metadata$Condition), 
                        MS1data$data,
                        stringsAsFactors = F)
names(MS1subset)[1] <- 'Condition'
MS1subset[, -1] <- data.frame(scale(MS1subset[, -1], center = train_scaling$center, scale = train_scaling$scale))

# project the controls
MS1ctrl <- MS1subset[MS1subset$Condition == 'ctrl', c('Condition', as.character(coefs$Metabo))]
MS1ctrl <- data.frame(as.matrix(MS1ctrl[, -1]) %*% pca$rotation )
MS1ctrl <- cbind(MS1ctrl[,1:2], row.names(MS1ctrl))
names(MS1ctrl)[3] <- 'Sample.ID'
MS1ctrl$Condition <- 'Control'

plotdata.ctrl <- rbind(plotdata, MS1ctrl)
pca.MS1.ctrl <- ggplot(plotdata.ctrl, aes(x=PC1, y=PC2, col = Condition, shape = Condition, label = Sample.ID, alpha = Condition)) + 
  geom_point(size = 2) + 
  xlab(paste0('PC1: explaining ',round(vars[1], digits = 3)*100, '% of the variance')) + 
  ylab(paste0('PC2: explaining ',round(vars[2], digits = 3)*100, '% of the variance')) + theme_bw()  +
  scale_shape_manual(values = c(15, 19, 17)) +
  scale_color_manual(values = c('black', '#3C568E', '#CC3333')) +
  scale_alpha_manual(values = c(1, .5, .5)) +
  ggtitle('Projected controls')

# project the transitions
MS1trans <- MS1subset[MS1subset$Condition == 'Transition', c('Condition', as.character(coefs$Metabo))]
MS1trans <- data.frame(as.matrix(MS1trans[, -1]) %*% pca$rotation )
MS1trans <- cbind(MS1trans[,1:2], row.names(MS1trans))
names(MS1trans)[3] <- 'Sample.ID'
MS1trans$Condition <- 'Transition'

plotdata.trans <- rbind(plotdata, MS1trans)
pca.MS1.trans <- ggplot(plotdata.trans, aes(x=PC1, y=PC2, col = Condition, shape = Condition, label = Sample.ID, alpha = Condition)) + 
  geom_point(size = 2) + 
  xlab(paste0('PC1: explaining ',round(vars[1], digits = 3)*100, '% of the variance')) + 
  ylab(paste0('PC2: explaining ',round(vars[2], digits = 3)*100, '% of the variance')) + theme_bw()  +
  scale_shape_manual(values = c(19, 17, 8)) +
  scale_color_manual(values = c('#3C568E', '#CC3333', 'black')) +
  scale_alpha_manual(values = c(.5, .5, 1)) +
  ggtitle('Projected transitioners')
```

```{r, include = T, fig.align = 'center', fig.width = 12, fig.height = 15}
adds <- theme(legend.position = 'none', 
              plot.title = element_text(size = 18, face = "bold"),
              axis.text = element_text(size = 12),
              axis.title = element_text(size = 12),
              plot.margin = unit(c(.5,.5,.5,.5), "cm"))
grid.arrange(pca.MS3 + adds, 
             pca.MS1 + adds, 
             pca.MS3.trans + adds,
             pca.MS1.trans + adds,
             pca.MS3.ctrl + adds,
             pca.MS1.ctrl + adds, ncol = 2)
```
**Figure 3.** Principal component analyses of the 31 selected metabolic features on the training (MS3) cohort and the test (MS1) cohort. The PMS patients have been visualized as *red triangles* and the RRMS patients as *blue points*.

```{r, include = T, fig.align = 'center', fig.width = 12, fig.height = 10}
grid.arrange(pca.MS3.age + adds, 
             pca.MS1.age + adds, 
             pca.MS3.sex + adds, 
             pca.MS1.sex + adds, ncol = 2)
```

# Clinical trial

```{r, include = T, fig.align = 'center', fig.width = 7, fig.height = 5}
bplot
```
**Figure 4.** Projected PC1 scores of the follow-up samples from the clinical trial of rituximab. Note, this plot does only include samples from the MS3 cohort.

\
**Mann-Whitney tests** between the time points in the clinical trial.

```{r}
plotdata.time$Time <- substrRight(plotdata.time$Sample.ID, 2)
plotdata.time$Time[-grep('T', plotdata.time$Time)] <- ''
plotdata.time$Subject <- substr(plotdata.time$Sample.ID, 1,4)
plotdata.time$Subject <- gsub('T', '', plotdata.time$Subject) 
```

```{r, include = T, echo = T}
indiv <- intersect(plotdata.time[plotdata.time$Time == 'T1', 'Subject'],
                   plotdata.time[plotdata.time$Time == 'T2', 'Subject'])
wilcox.test(plotdata.time[paste0(indiv, 'T1'), 'PC1'], 
            plotdata.time[paste0(indiv, 'T2'), 'PC1'], paired = T)
```

```{r, include = T, echo = T}
indiv <- intersect(plotdata.time[plotdata.time$Time == 'T1', 'Subject'],
                   plotdata.time[plotdata.time$Time == 'T3', 'Subject'])
wilcox.test(plotdata.time[paste0(indiv, 'T1'), 'PC1'], 
            plotdata.time[paste0(indiv, 'T3'), 'PC1'], paired = T)
```

```{r, include = T, echo = T}
indiv <- intersect(plotdata.time[plotdata.time$Time == 'T1', 'Subject'],
                   plotdata.time[plotdata.time$Time == 'T4', 'Subject'])
wilcox.test(plotdata.time[paste0(indiv, 'T1'), 'PC1'], 
            plotdata.time[paste0(indiv, 'T4'), 'PC1'], paired = T)
```

# Summarize selection

```{r}
selected <- as.character(coefs[sort(abs(coefs$Coefficient), 
                                    decreasing = T, index.return = T)$ix, 'Metabo'])
row.names(featureMatch) <- paste0('X', featureMatch$MS3)
selected <- featureMatch[selected, ]

selected$mz <- round(MS3data$untouched[selected$MS3, 'mz_cf'], digits = 2)
selected$rt <- round(MS3data$untouched[selected$MS3, 'rt_cf'], digits = 2)

# extract untouched data for MS3 training set
MS3dat <- data.frame(cbind(MS3meta[, c('Condition', 'Age', 'Gender')], 
                           data.frame(t(MS3data$untouched[selected$MS3, row.names(MS3meta)]))), stringsAsFactors = F)

# fit models
selected$FC.MS3 <- NA
selected$pVal.MS3 <- NA
selected$sex.MS3 <- NA
for (i in 4:ncol(MS3dat)) {
  input <- MS3dat[, c(1:3, i)]
  names(input)[4] <- 'Metabo'
  
  input$Condition <- factor(input$Condition, levels = c('PMS', 'RRMS'))
  lm_model <- lm(Metabo ~ Condition + Gender, data = input)
  em <- emmeans(lm_model, 'Condition')
  ctr <- data.frame(emmeans::contrast(em, "pairwise"))
  
  selected$FC.MS3[i-3] <- round(ctr$estimate, digits = 2)
  selected$pVal.MS3[i-3] <- ctr$p.value
  
  em <- emmeans(lm_model, 'Gender')
  ctr <- data.frame(emmeans::contrast(em, "pairwise"))
  
  selected$sex.MS3[i-3] <- ctr$p.value
}
selected$pVal.MS3 <- stars.pval(selected$pVal.MS3)
selected$sex.MS3 <- stars.pval(selected$sex.MS3)

# extract untouched data for MS1 testing set
MS1dat <- data.frame(cbind(MS1meta[, c('Condition', 'Age', 'Gender')],  
                           data.frame(t(MS1data$untouched[selected$MS1, row.names(MS1meta)]))), stringsAsFactors = F)
names(MS1dat)[1] <- 'Condition'

# fit models
selected$FC.MS1 <- NA
selected$pVal.MS1 <- NA
selected$sex.MS1 <- NA
for (i in 4:ncol(MS1dat)) {
  input <- MS1dat[, c(1:3, i)]
  names(input)[4] <- 'Metabo'
  
  input$Condition <- factor(input$Condition, levels = c('PMS', 'RRMS'))
  lm_model <- lm(Metabo ~ Condition + Gender, data = input)
  em <- emmeans(lm_model, 'Condition')
  ctr <- data.frame(emmeans::contrast(em, "pairwise"))
  
  selected$FC.MS1[i-3] <- round(ctr$estimate, digits = 2)
  selected$pVal.MS1[i-3] <- ctr$p.value
  
  em <- emmeans(lm_model, 'Gender')
  ctr <- data.frame(emmeans::contrast(em, "pairwise"))
  
  selected$sex.MS1[i-3] <- ctr$p.value
}
selected$pVal.MS1 <- stars.pval(selected$pVal.MS1)
selected$sex.MS1 <- stars.pval(selected$sex.MS1)

write.table(selected, file = 'table_of_selection.csv', sep = ';')
```

```{r, include = T}
# find inconsistent features
selected$inconsistent <- as.numeric(sign(selected$FC.MS3) != sign(selected$FC.MS1))
selected$inconsistent[selected$inconsistent == 0] <- ''

# create table
kable(selected, caption = 'Summarized data for the selected metabolic features using emmeans.', 
      row.names = F, align = rep('c', ncol(selected))) 
```

# Boxplots

```{r, include = T, fig.align = 'center', fig.width = 13, fig.height = 12}
# extract untouched MS3 data
MS3dat <- data.frame(cbind(as.character(MS3data$metadata$Condition), 
                           as.character(MS3data$metadata$Collection),
                           data.frame(t(MS3data$untouched[selected$MS3, -c(1,2)]))), stringsAsFactors = F)
names(MS3dat)[1:2] <- c('Condition', 'Collection')

# center
MS3dat[, paste0('X', selected$MS3)] <- scale(MS3dat[, paste0('X', selected$MS3)], center = T, scale = F)

MS3dat <- MS3dat[MS3dat$Collection == 1, ]
MS3dat$Cohort <- 'MS3'

# extract untouched MS1 data
MS1dat <- data.frame(cbind(as.character(MS1data$metadata$Condition), 
                           data.frame(t(MS1data$untouched[selected$MS1, -c(1,2)]))), stringsAsFactors = F)
names(MS1dat) <- c('Condition', paste0('X', selected$MS3))
MS1dat$Cohort <- 'MS1'

# center 
MS1dat[, paste0('X', selected$MS3)] <- scale(MS1dat[, paste0('X', selected$MS3)], center = T, scale = F)

comb <- rbind(MS3dat[, c('Cohort', 'Condition', paste0('X', selected$MS3))],
              MS1dat[, c('Cohort', 'Condition', paste0('X', selected$MS3))])


comb <- gather(comb, 'Metabolite', 'Value', -c(Condition, Cohort))
comb$Group <- paste(comb$Cohort, comb$Condition)
comb <- comb[-which(comb$Condition == 'Transition'), ]
comb <- comb[-which(is.na(comb$Value)),]

comb$Metabolite <- factor(comb$Metabolite, levels = paste0('X', selected$MS3))
ggplot(comb, aes(Group, Value, fill = Group)) + 
  geom_boxplot() + scale_fill_manual(values = c(brewer.pal(3, 'Reds'), brewer.pal(3, 'Blues'))) +
  facet_wrap(~Metabolite, scales = 'free') + 
  ylab('log2 intensity') + xlab('') + theme_bw() + 
  theme(axis.text.x = element_blank(), 
        axis.title = element_text(size = 12), 
        axis.text = element_text(size = 12))
```